Experience Report: Haskell in Computational Biology

### 2012-07-01 mrfy_experience_report.tex

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
\remark{This looks wrong.  $x$ should be subscripted with a
query-sequence index, but $\pairedwith j$ is a model index.}
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.

@@

But \mrfy's job is not to run Viterbi's algorithm on large models;
\mrfy's job is to detect homologies to structures on which Viterbi's
algorithm is intractable.
We~benchmarked \mrfy\ 
using a model of an ``8-bladed beta propeller,''
which has 343 nodes, of which NNN appear in 40~beta strands.
The models between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
CAN WE PLEASE REPLACE THE TEXT ABOVE WITH A DESCRIPTION OF WHAT'S
GOING ON IN THE CODE WE ACTUALLY RAN???
On~these smaller problems, 
performance is good enough that \mrfy\ can do its job.
%%
%%
%%
SUDDEN CHANGE OF BENCHMARK CONSIDERED HARMFUL.
%%
%%
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.

Because \mrfy\ works on many subproblems independently, we~found it easy
to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
Using
\texttt{Control.Parallel}, parallelizing this computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:efficiency} shows parallel efficiency when
using a model of an ``8-bladed beta propeller;''
the model has 343 ordinary nodes and 40~beta strands TOTALLING HOW
MANY NODES?
We~benchmarked \mrfy\ using a query sequence of
592 amino acids.
Figure~\ref{fig:speedup} SHOWS NOTHING OF VALUE BECAUSE RANDOMNESS
BREAKS EVERYTHING.  AND IT SHOULD HAVE ERROR BARS.
(Parallel efficiency is 1.0 when the speedup is equal to the number of cores
used.
Lower speedups mean lower efficiency.)

@@

We~can also make a preliminary comparison of performance.
SMURF can parse a simple protein structure, optimally, in under 10~seconds.
\remark{At this stage ``parse'' is kind of a new word.}
It~can parse some more complex structures in 2--3~minutes.
But when asked to parse the most complex structures, it runs out of
memory and has to be killed, even on a system with 128GB of~RAM.
By~contrast, MRFy can parse any of these structures, approximately,
in 30~seconds.
That's fast enough
%This performance
%meets our goals, and it means that MRFy can be used
to scan whole genomes.

@@

\bibliographystyle{plainnat}

>>>>

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith i}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith i}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.

@@

But \mrfy's job is not to run Viterbi's algorithm on large models;
\mrfy's job is to detect homologies to structures on which Viterbi's
algorithm is intractable.
We~benchmarked \mrfy\ 
using a model of an ``8-bladed beta propeller,''
which has 343 nodes, of which 178 appear in 40~beta strands.
The models between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
CAN WE PLEASE REPLACE THE TEXT ABOVE WITH A DESCRIPTION OF WHAT'S
GOING ON IN THE CODE WE ACTUALLY RAN???
On~these smaller problems, 
performance is good enough that \mrfy\ can do its job.
%%
%%
%%
SUDDEN CHANGE OF BENCHMARK CONSIDERED HARMFUL.
%%
%%
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.

Because \mrfy\ works on many subproblems independently, we~found it easy
to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
Using
\texttt{Control.Parallel}, parallelizing this computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:efficiency} shows parallel efficiency when
using a model of an ``8-bladed beta propeller;''
the model has 165 ordinary nodes and 40~beta strands totalling
178 nodes.
We~benchmarked \mrfy\ using a query sequence of
592 amino acids.
Figure~\ref{fig:speedup} SHOWS NOTHING OF VALUE BECAUSE RANDOMNESS
BREAKS EVERYTHING.  AND IT SHOULD HAVE ERROR BARS.
(Parallel efficiency is 1.0 when the speedup is equal to the number of cores
used.
Lower speedups mean lower efficiency.)

@@

@@

\bibliographystyle{abbrvnat}
